---
title: "mini-data-analysis-milestone1"
author: "Lorenzo Lindo"
date: "07/10/2021"
output: github_document
---

# STAT545A - Mini Data Analysis Project: Milestone 1 (LINDO, Lorenzo)

## Load the required packages.
Run the following code to load the required packages for this mini data analysis.
```{r}
library(datateachr)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

## Task 1 - Exploring My Datasets of Interest
### These datasets that will be used explored for this mini data analysis.
Dataset Number | Dataset Name | Description of Data Set | Credit | Dimensions
---------------|--------------|-------------------------|--------|---------
1 | cancer_sample | Diagnostic Breast Cancer Data | Acquired courtesy of UCI Machine Learning Repository | 569 rows and 32 columns
2 | steam_games | Contains games from Steam shop with detailed data downloaded from kaggle with a CC0 license | Acquired courtesy of Kaggle | 40833 rows and 21 columns
3 | building_permits | Data about the City of Vancouver's building permits | Acquired courtesy of the City of Vancouver's Open Data Portal | 20680 rows and 14 columns
4 | vancouver_trees | Data about Vancouver's trees | Acquired courtesy of the City of Vancouver's Open Data Portal | 146611 rows and 20 columns

#### Exploring the cancer_sample dataset
```{r}
head(cancer_sample) #Returns the first few rows
dim(cancer_sample) #Returns the dimensions of the dataset (how many rows and columns)
class(cancer_sample) #Returns the class of the dataset
cancer_sample %>% 
  group_by(diagnosis) %>% 
  summarise(perimeter_mean_median = mean(perimeter_mean, na.rm = TRUE))
```

#### Exploring the steam_games dataset
```{r}
head(steam_games) #Returns the first few rows
dim(steam_games) #Returns the dimensions of the dataset (how many rows and columns)
class(steam_games) #Returns the class of the dataset
steam_games %>% 
  group_by(genre) %>% 
  summarize(n())
```

#### Exploring the building_permits dataset
```{r}
head(building_permits) #Returns the first few rows
dim(building_permits) #Returns the dimensions of the dataset (how many rows and columns)
class(building_permits) #Returns the class of the dataset
building_permits %>% 
  group_by(type_of_work) %>% 
  summarize(n())
```

#### Exploring the vancouver_trees dataset
```{r}
head(vancouver_trees) #Returns the first few rows
dim(vancouver_trees) #Returns the dimensions of the dataset (how many rows and columns)
class(vancouver_trees) #Returns the class of the dataset
vancouver_trees %>% 
  group_by(genus_name) %>% 
  summarize(n())
```

### Narrowing down to 2 datasets of interest
Dataset | Why I chose this?
--------|------------------
cancer_sample | I chose to continue exploring this dataset because my research is focussed on cancer and I wanted to practice data wrangling and analysis on a dataset related to cancer.
vancouver_trees | I chose to continue exploring this dataset because I am a born and raised Vancouverite and I really love nature, so I thought this might be very interesting.

### The final decision - which one will I continue analyzing?
Dataset | Possible Research Questions
--------|----------------------------
cancer_sample | How does the median of the mean area of nuclei compare between benign and malignant breast masses?
vancouver_trees | Is there a statistically significant difference in the number of trees of different genera?

**My Final Choice**: I will analyze the **vancouver_trees** dataset.  This will be the dataset that I will continue to explore.  While the cancer_sample dataset is interesting because my research focusses on cancer, this dataset only contains one categorical variable and all other variables are numerical.  Since there is only one categorical variable, it made it difficult for me to think of research questions besides asking if there is any difference between malignant and benign breast masses.  I also thought it would be fun to explore the diversity of trees in my hometown of Vancouver.

## Task 2 - Exploring the vancouver_trees dataset
### Some Initial Manipulation of the Datasets

#### Exercise Option 1: 
Plotting the distribution of the diameter of trees.  **Rationale**: This was done to get a sense of the distribution of the diameter variable.  The following code takes the vancouver_trees dataset and generates a histogram with diameter on the x axis and count on the y axis.  This histogram plots the distribution of the diameter of trees in this dataset.
```{r}
vancouver_trees %>% 
  ggplot(aes(x = diameter)) +
  geom_histogram()
```

#### Exercise Option 2: 
Calculating the expected circumference of trees based on the observed diameter and assigning it as a new variable.  **Rationale**: This was done to calculate what the expected tree circumference would be given the observed diameter of each tree observed.  The following code creates a new variable that contains the expected circumference based on the observed diameter of trees.  It calculates circumference by multiplaying the diameter by pi.  It also relocates both the diameter and expected circumference columns such that they appear after the std_street column.
```{r}
vancouver_trees %>% 
  mutate(expected_circumference = (diameter) * pi) %>% 
  relocate(diameter, .after = std_street) %>% 
  relocate(expected_circumference, .after = diameter)
```

#### Exercise Option 4: 
Exploring the relationship between the diameter of the trees and the latitude.  **Rationale**: This was an exploratory analysis to see if there is any trend between the diameter and the latitude.  If you run the following code, it doesn't seem like there is any trend on first glance.  The following code generates a scatterplot with diameter on the x axis and latitude on the y axis.
```{r}
vancouver_trees %>% 
  ggplot(aes(x = diameter, y = latitude)) +
  geom_point()
```
  
#### Exercise Option 6: 
A boxplot looking at the height range of Vancouver trees based on whether or not they have a root barrier installed.  **Rationale**: This was done to see if there is any difference in the height range of trees among those trees with or without a root barrier installed.  The following code first groups the data by the presence or absence of a root barrier, then generates a box plot.  The box plot contains the presence or absence (Y/N) of a root barrier on the x-axis, and the height range on the y-axis.  It colours the box plots based on the presence or absence of a root barrier.  I have also renamed the axes titles and added a plot title.
```{r}
vancouver_trees %>% 
  group_by(root_barrier) %>% 
  ggplot(aes(x = root_barrier, y = height_range_id, colour = root_barrier)) +
  geom_boxplot() +
  labs(title = "Frequency of Root Barriers Among Vancouver Trees" ) +
  xlab("Root Barrier? (variable = root_barrier)") +
  ylab("Height Range (variable = height_range_id") +
  theme_minimal()
```  

## Task 3 - Writing my Research Questions
Below are my proposed research questions for this dataset.  I also include some rationale and expected outcomes

Question Number | Research Question
----------------|------------------
1 | Is there a significant difference in the number of trees between the East and West sides of Vancouver? I would be interested in seeing if there is any difference in the frequency of trees between the East and West sides and if this difference is statistically significant.  I would need to create a new categorical variable to classify trees as either in the East or West sides (I could divide Vancouver based on longitude) then plot a graph.
2 | Does the presence of a root barrier affect the diameter of trees? Root barriers are structures that are meant to guide tree roots down and away from human-made structures (e.g. building foundations, gas pipes, electrical wires).  I am interested to see if these root guides have any affect on the diameter of trees.
3 | Is there a difference in the frequency of tree species within a specific genus in a neighbourhood of interest? Within a specified neighbourhood (e.g. Riley Park), and within a given genus of interest (e.g. Prunus), are there any differences in the frequency among tree species?
4 | Are there any differences in the height range of all trees between neighbourhoods in Vancouver?  Neighbourhoods have a lot of differences, including local climates, levels of urbanization, and proximity to bodies of water (among others).  I would like to see if there is any statistically significant difference in the mean number of trees among Vancouver's neighbourhoods.
