---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome back to your mini data analysis project!

This time, we will explore more in depth the concept of *tidy data*, and hopefully investigate further into your research questions that you defined in milestone 1.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. Is there a significant difference in the number of trees between the different sides of the street (EVEN vs. ODD) in Vancouver? I would be interested in seeing if there is any difference in the frequency of trees between the EVEN and ODD sides of Vancouver streets and if this difference is statistically significant.  NOTE - I had to modify this question from Milestone 1 as there are not enough decimals in the longitude to differentiate between East or West sides.
2. Does the presence of a root barrier affect the diameter of trees? Root barriers are structures that are meant to guide tree roots down and away from human-made structures (e.g. building foundations, gas pipes, electrical wires).  I am interested to see if these root guides have any affect on the diameter of trees.
3. Is there a difference in the frequency of tree species within a specific genus in a neighbourhood of interest? Within a specified neighbourhood (e.g. Riley Park), and within a given genus of interest (e.g. Prunus), are there any differences in the frequency among tree species?
4. Are there any differences in the height range of all trees between neighbourhoods in Vancouver?  Neighbourhoods have a lot of differences, including local climates, levels of urbanization, and proximity to bodies of water (among others).  I would like to see if there is any statistically significant difference in the mean number of trees among Vancouver's neighbourhoods.
<!----------------------------------------------------------------------------->

### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
### Research Question 1
#### SUMMARIZING: Compute the number of observations for at least one of your categorical variables: species_name and genus_name
The code below will return the number unique species represented in the dataset.  This doesn't really help me answer my research question, but it does give me a sense of the taxonomic diversity of this dataset.
```{r}
unique_species <- vancouver_trees %>% group_by(species_name) %>% summarize(count = n())
print(paste0("The number of unique species in the dataset is ", nrow(unique_species)))

unique_genera <- vancouver_trees %>% group_by(genus_name) %>% summarize(count = n())
print(paste0("The number of unique genera in the dataset is ", nrow(unique_genera)))

```
The code below will return the number of observations where the species name is "Americana".  This also doesn't really help me answer the research question, but does give me some practice quantifying the frequency of a given value for a given variable.
```{r}
num_Americana <- sum(with(vancouver_trees, species_name == "AMERICANA"))
print(paste0("The number of observations where the species name is 'AMERICANA' is ", num_Americana))

```

#### GRAPHING: The following code will display the number of trees found in the EVEN side of the streets compared to the ODD side of the streets in Vancouver then graph it with a logarithmic y-axis.  The helps me answer this research question by allowing me to compare the number of trees that appear on the EVEN or ODD sides of streets in Vancouver.
```{r}
even_vs_odd <- vancouver_trees %>% select(street_side_name) %>%  summarize(num_odd = sum(street_side_name == "ODD"), num_even = sum(street_side_name == "EVEN"))
print(even_vs_odd)

vancouver_trees %>% filter(street_side_name %in% c("EVEN", "ODD")) %>% ggplot(aes(x = street_side_name)) + 
  geom_bar() +
  scale_y_continuous(trans = "log10") +
  xlab("Even or Odd Side of Street") +
  ylab("Number of Trees") +
  ggtitle("Number of trees found in Even or Odd sides of the street in Vancouver") +
  theme_minimal()

```

### Research Question 2

#### SUMMARIZING: Compute the number of observations for at least one of your categorical variables.
The following code allows me to quantify how many trees do and do not have a root barrier.  This helps me with the research question as I will be comparing the diameter of trees with and without root barriers.
```{r}
no_root_barrier <- sum(with(vancouver_trees, root_barrier == "N"))
has_root_barrier <- sum(with(vancouver_trees, root_barrier == "Y"))
print(paste0("The number of trees that have a root barrier is ", has_root_barrier, " and the number of trees that do not have a root barier is ", no_root_barrier))
```

#### GRAPHING: Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
The code below will generate a bar graph comparing the number of trees with and without a root barrier.  This also helps me answer the research question by giving me a visual sense of how many trees do and do not have a root barrier.
```{r}
vancouver_trees %>% filter(root_barrier %in% c("N", "Y")) %>% ggplot(aes(x = root_barrier)) +
  geom_bar() +
  scale_y_continuous(trans = "log10") +
  xlab("Absence (N) or Presence (Y) of a Root Barrier") +
  ylab("Number (Count)") +
  ggtitle("Quantification of Trees With or Without a Root Barrier") +
  theme_minimal()

```

### Research Question 3

#### SUMMARIZING: Based on two categorical variables, calculate two summary statistics of your choosing. 
The code below will calculate the mean and range of the frequency of ULMUS AMERICANA among neighbourhoods in Vancouver.  It will first create a tibble that returns the number of ULMUS AMERICANA per neighbourhood.  This code helps me build a pipeline for calculating summary statistics that will help me answer this research question.
```{r}
freq_ulmus_americana <- vancouver_trees %>% 
  filter(species_name == "AMERICANA") %>% 
  group_by(neighbourhood_name) %>% 
  summarise(americana_per_neighbourhood = n())

print(freq_ulmus_americana)

mean_americana <- freq_ulmus_americana %>% summarize(mean = mean(americana_per_neighbourhood))
print(paste0("The mean number of Ulmus Americana among Vancouver's neighbourhoods is ", mean_americana))

range_americana <- freq_ulmus_americana %>% summarize(range = range(americana_per_neighbourhood))
print(paste0("The range of the frequency of Ulmus Americana among Vancouver's neighbourhoods is ", range_americana))

```

#### GRAPHING: Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
The code below will make a graph that shows the distribution of the frequency of Ulmus Americana among Vancouver's neighbourhoods.  This is also provides a pipeline that I can use for further analyses for this research question.
```{r}
vancouver_trees %>% filter(species_name == "AMERICANA") %>% ggplot(aes(neighbourhood_name)) +
  geom_histogram(stat = "count") +
  ggtitle("Number of Ulmus Americana in Vancouver Neighbourhoods") +
  ylab("Number of Ulmus Americana") +
  xlab("Neighbourhood Name") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(trans = "log10")

```


### Research Question 4

#### SUMMARIZING: Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
The code below will calculate the mean of the heights of trees represented in this dataset.
```{r}
height_mean <- vancouver_trees %>% summarize(mean_height = mean(height_range_id))
print(height_mean)
```

The code below will return the range of the heights of trees represented in this dataset.
```{r}
height_range <- vancouver_trees %>% summarize(height_range = range(height_range_id))
print(height_range)

```
The code below will calculate the standard deviation of the height of trees in this dataset.
```{r}
height_sd <- sd(vancouver_trees$height_range_id)
print(height_sd)
```

The code below will calculate the standard error of the height of trees in this dataset.
```{r}
height_se <- sd(vancouver_trees$height_range_id) / sqrt(nrow(vancouver_trees))
print(height_se)

```

#### GRAPHING: Create a graph out of summarized variables that has at least two geom layers.
The code below will create a bar graph that shows the count of trees present in each neighbourhood in Vancouver.  I will be comparing the height range of trees in Vancouver neighbourhoods for this research question.  This graph will give me a sense of the frequency of trees in each Vancouver neighbourhood.  Since neighbourhoods vary considerably with regards to the number of trees they contain, it will be a caveat during comparisons as there is a different sample size of trees from each neighbourhood.
```{r}
unique_neighbourhoods <- vancouver_trees %>% group_by(neighbourhood_name) %>% summarize(num = n())

unique_neighbourhoods %>% ggplot(aes(x = neighbourhood_name, y = num)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  ggtitle("Number of Trees in Vancouver Neighbourhoods") +
  ylab("Count (Number)") +
  xlab("Neighbourhood Name")

```


<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
**RESEARCH QUESTION 1** From Milestone 1, I initially wanted to investigate the frequency of trees in the EAST vs. WEST sides of Vancouver.  Unfortunately, the longitude data is not accurate enough (not enough decimal points) to differentiate between EAST and WEST sides.  As such, I modified the research question to look at the frequency of trees between the different sides of the street - EVEN vs. ODD.  This question is pretty much answered and it looks like it is very close - similar numbers of trees on EVEN vs. ODD sides of streets in Vancouver.

**RESEARCH QUESTION 2** I am getting closer towards answering this research question.  In Milestone 2, I have now quantified the number of trees with and without root barriers.  What I will have to do next is plot the distributions of tree diameter among each group (i.e. with and without root barriers).  I could then run statistical analyses to compare some summary statistics, such as the mean.

**RESEARCH QUESTION 3** In this Milestone, I have created a pipeline to determine the frequency of a given tree species among neighbourhoods in Vancouver.  Next steps could include querying other species of interest within the genus I selected, or I could even look at other genera.  What could be interesting for future analyses would be to compare the diversity between neighbourhoods.  For example, I could quantify the mean number of species within

**RESEARCH QUESTION 4** In this Milestone, I computed some summary statistics, including mean and range, that describe the distribution of the height of trees in this dataset.  For future analyses in this research question, I could compute the mean height within each neighbourhood and then plot the distribution to get an idea of how each neighbourhood compares.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
As my dataset has >8 variables, I will pick 8 variables for this question.  I will pick the first 8 columns and assign it to a new variable name.
```{r}
trees_subset <- vancouver_trees %>% select(1:8)
head(trees_subset)
dim(trees_subset)
```
Let's see if this subset of vancouver_trees satisfies the definition of *tidy* data as stated above:

- [x] Each row is an **observation**
- [x] Each column is a **variable**
- [] Each cell is a **value**

As we can see, there are some NA entries in our dataset - and our dataset is therefore NOT TIDY.

If I wanted to increase the rigor of my assessment of the "tidiness" of this dataset, I could do the following to quantify how many NA values are present for each variable:
```{r}
num_NA_tree_id <- sum(is.na(trees_subset$tree_id))
print(num_NA_tree_id)

num_NA_civic_number <- sum(is.na(trees_subset$civic_number))
print(num_NA_civic_number)

num_NA_std_street <- sum(is.na(trees_subset$std_street))
print(num_NA_std_street)

num_NA_genus_name <- sum(is.na(trees_subset$genus_name))
print(num_NA_genus_name)

num_NA_species_name <- sum(is.na(trees_subset$species_name))
print(num_NA_species_name)

num_NA_cultivar_name <- sum(is.na(trees_subset$cultivar_name))
print(num_NA_cultivar_name)

num_NA_common_name <- sum(is.na(trees_subset$common_name))
print(num_NA_common_name)

num_NA_assigned <- sum(is.na(trees_subset$assigned))
print(num_NA_assigned)


```

<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->
As I have shown above, this subset (trees_subset) of the vancouver_trees dataset is NOT tidy because there are many missing values - particularly for the cultivar_name variable.

In order to tidy this dataset, I will remove all the entries with NA values for the cultivar.

```{r}
trees_subset <- vancouver_trees %>% select(1:8)
head(trees_subset)
dim(trees_subset)

trees_subset_removeNA <- na.omit(trees_subset)
head(trees_subset_removeNA)
dim(trees_subset_removeNA)

print(paste0("The number of observations in the original datasett is ", nrow(trees_subset), " whereas the number of rows in the tidy dataset after removing NA values is ", nrow(trees_subset_removeNA)))
```

As you can see, we have gone from 146611 rows (observations) in trees_subset to 79063 rows (observations) in trees_subset_removeNA.  This means that there were 67559 rows containing an NA value.  This is concordant with the number of entries that have an NA for cultivar_name that we quantified in the section above.

That being said, besides the NA values for the cultivar_name, the dataset is otherwise tidy because each column is a variable and each row is an observation.  Since it is tidy, we can untidy the data.  For example, we can combine the genus_name and species_name to give the scientific_name for each tree and therefore make the data untidy.  This makes the data untidy since we can no longer run analyses looking at genera or species separately.  
```{r}
trees_scientific_names <- trees_subset %>% unite(col = "scientific_name", genus_name, species_name, sep = " ")
head(trees_scientific_names)
```
Then, to tidy the data, we can separate the scientific name into its genus_name and species_name.
```{r}
trees_restored_names <- trees_scientific_names %>% separate(col = scientific_name, into = c("genus_name", "species_name"))
print(trees_restored_names)

```

<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelvant columns, etc.). 

<!--------------------------- Start your work below --------------------------->
Based on the analyses that I have completed in this Milestone (Milestone 2), I would look to continue pursing **RESEARCH QUESTION 2** and **RESEARCH QUESTION 4**.  I have chosen to pursue **Research Question 2** because I am generally curious about how root barriers can affect the growth of trees.  I believe that this could give some insight into the environmental impact that human-made structures can have on trees.  I also think it would be an interesting opportunity to attempt statistical analyses in R.  I have chosen to pursue **Research Question 4** because I am interested in seeing how neighbourhoods compare to one another - the local biome of each neighbourhood is different from one another, so it would be intriuging to see how tree height compares.

To facilitate my analyses as part of Research Questions 2 and 4, I will create a version of my data that is appropriate.  The first function will select those columns that I am interested in looking at.  While my analyses will primarily look at the tree_id, genus_name, species_name, root_barrier, neighbourhood_name, height_range_id, and diameter, I will also include the date_planted in case I wanted to do some other exploratory anaylses looking at how things trend with time.  In order to simplify my analyses, I will separate the date into year, month, and day.  Thirdly, since I would like to focus my analyses on diamter and height, I will remove those entries which have NA values for those variables. Fourthly, to make it easier when coding, I will rename some of the variables to simplify calling them.
```{r}
tidy_vancouver_trees <- vancouver_trees %>% select(tree_id, genus_name, species_name, root_barrier, neighbourhood_name, height_range_id, diameter, date_planted) %>%
  separate(date_planted, into = c("year", "month", "day"), sep = "-") %>% 
  filter(!is.na(height_range_id)) %>% 
  filter(!is.na(diameter)) %>% 
  rename(tree = tree_id, genus = genus_name, species = species_name, neighbourhood = neighbourhood_name, height = height_range_id)

print(tidy_vancouver_trees)
dim(tidy_vancouver_trees)

```

<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 